Nested Sampling with Multiple Objective Functions

### 2018-06-16 doc/ms.tex
A few edits

\section{Example 1: Known solution}
In the software repository, this example is implemented in
{\tt examples/Demo.h} and {\tt examples/Demo.cpp}.
We tested the algorithm on a simple 100-dimensional example, where the
prior is a uniform distribution between 0 and 1 for each coordinate:
\begin{eqnarray}
\pi(\xx) &=& \prod_{i=1}^{100}
\left\{
\begin{array}{lr}
1, & x_i \in [0, 1]\\
0, & \textnormal{otherwise.}
\end{array}
\right.
\end{eqnarray}
The first objective function is quadratic in form, such that the canonical
distributions are iid normal centered at $(0.5, 0.5, ..., 0.5)$:
\begin{eqnarray}
E_1(\xx) &=& \sum_{i=1}^{100} \left(\frac{x_i - 0.5}{0.1}\right)^2.
\end{eqnarray}
The second objective function, given by
\begin{eqnarr \sum_{i=1}^{100}\sin^2(10 \pi x_i),
\end{eqnarray}
upweights states where coordinates $\{x_i\}$
are close to 0, 0.1, 0.2, ..., 0.9 and 1.
Since both of the objective functions are simple sums over the coordinates,
the canonical distributions are products of independent distributions for
each coordinate.

@@

\section{Example 3: ``MaxEnt'' image deconvolution example}

Problems with more than one objective function can also arise in data
analysis. Consider, for example, the problem of deblurring a noisy
image. A common approach to these problems is regularized inversion,
such as the ``maximum entropy'' technique of \citet{gull}. It is important
to distinguish between this method and the maximum entropy principle used
to update or assign probabilities \citep{caticha}.

The unknown parameters are a set of pixel intensities
$\{x_i\}_{i=1}^n$ where $n$ is the number of pixels and $x_i \geq 0$.
The dataset is also a collection of pixel intensities $\{D_i\}_{i=1}^n$.
The prior for the dataset given the parameters is a normal distribution
with mean given by the underlying image $\{xi\}$ convolved by
a point spread function. The standard deviation of each pixel is
$\sigma$, assumed known. This is usually written as

\begin{eqnarray}
\boldsymbol{D} = \boldsymbol{M}\boldsymbol{x} + \boldsymbol{n}
\end{eqnarray}
where $\boldsymbol{n}$ is the noise in each pixel and $\boldsymbol{M}$ is
the blurring operator written in matrix form.

The prior for the pixel intensities $\{x_i\}$ has an ``entropic'' form,
where images are upweighted (relative to a uniform prior) if the pixel
intensities have a low dynamic range. The prior density is defined conditional
on a hyperparameter $\lambda$:
\begin{eqnarray}
p(\boldsymbol{x} | \lambda) &=&
\frac{\exp\left[\lambda S(\xx)\right]}{\int \exp\left[\lambda S(\xx)\right] \, d\xx}
\label{eq:entropy_prior}
\end{eqnarray}
where
\begin{eqnarray}
S(\xx) &=& -\sum_{i=1}^n x_i \ln x_i
\end{eqnarray}
is the ``configurational entropy'' of the image (crucially, this is a function of
an image, and not a function of a probability distribution).
The denominator in Equation~\ref{eq:entropy_prior} ensures the conditional prior
is normalized.

In the 1980s, this prior was thought to have a fundamental status, but this
is no longer the case \citep{massinf}, although the function $S(\xx)$ does
have special properties when its argument is a measure \citep{knuth_skilling}.
However, it is still a prior whose
consequences we might be interested in.

Instead, we can consider the configurational
entropy $S(\xx)$ as simply a function of an image that may
be relevant somehow, and that we are uncertain about. The joint posterior
for $\xx$ and $\lambda$ is:
\begin{eqnarray}
p(\lambda, \xx | \data) &=&
\frac{p(\lambda)p(\xx | \lambda)p(\data | \xx)}
{p(\data)}\\
&=&
\frac{p(\lambda)p(\xx | \lambda)p(\data | \xx)}
{p(\data)}
\end{eqnarray}

The posterior for the underlying image given the data is
\begin{eqnarray}
p(\boldsymbol{x} | \boldsymbol{D}, \lambda)
&=& \frac{p(\boldsymbol{x} | \lambda)p(\boldsymbol{D}|\boldsymbol{x}, \lambda)}
{\int p(\boldsymbol{x}|\lambda)p(\boldsymbol{D}|\boldsymbol{x}, \lambda) \,d^n\boldsymbol{m}}\\
&=& \frac{\exp\left[-\lambda \sum_i x_i\ln x_i
-\frac{1}{2\sigma^2}\left(D_i - \sum_{j} P_{ij}x_{j}\right)^2\right]}
{\int \exp\left[-\lambda \sum_i x_i\ln x_i
-\frac{1}{2\sigma^2}\left(D_i - \sum_{j} P_{ij}x_{j}\right)^2\right]\, d^n\boldsymbol{m}}
\end{eqnarray}

This is a `doubly intractable' problem \citep{murray}, since an integral, whose
result is a function of the parameters, appears in the {\em denominator} of the
target density. However, in our case the integral in the
denominator arises from the prior for the parameters, rather than from the
likelihood function.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Conclusions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{BJB acknowledges research and study leave funding from the
University of Auckland. We would like to thank Gabor Csanyi, Livia B. Partay,
Rob Salomone, John Skilling, etc etc.}

>>>>

\section{Example 1: Known solution}
In the software repository, this example is implemented in
{\tt examples/Demo.h} and {\tt examples/Demo.cpp}.
We tested the algorithm on a simple 100-dimensional example, where the
prior is a uniform distribution between 0 and 1 for each coordinate:
\begin{eqnarray}
\pi(\xx) &=& \prod_{i=1}^{100}
\left\{
\begin{array}{lr}
1, & x_i \in [0, 1]\\
0, & \textnormal{otherwise.}
\end{array}
\right.
\end{eqnarray}
The first objective function is quadratic in form, such that the canonical
distributions are iid normal centered at $(0.5, 0.5, ..., 0.5)$:
\begin{eqnarray}
E_1(\xx) &=& \sum_{i=1}^{100} \left(\frac{x_i - 0.5}{0.1}\right)^2.
\end{eqnarray}
The second objective function, given by
\begin{eqnarray}
E_2(\xx) &=& \sum_{i=1}^{100}\sin^2(10 \pi x_i),
\end{eqnarray}
upweights states where coordinates $\{x_i\}$
are close to 0, 0.1, 0.2, ..., 0.9 and 1.
Since both of the objective functions are simple sums over the coordinates,
the canonical distributions are products of independent distributions for
each coordinate.

@@

%\section{Example 3: ``MaxEnt'' image deconvolution example}

%Problems with more than one objective function can also arise in data
%analysis. Consider, for example, the problem of deblurring a noisy
%image. A common approach to these problems is regularized inversion,
%such as the ``maximum entropy'' technique of \citet{gull}. It is important
%to distinguish between this method and the maximum entropy principle used
%to update or assign probabilities \citep{caticha}.

%The unknown parameters are a set of pixel intensities
%$\{x_i\}_{i=1}^n$ where $n$ is the number of pixels and $x_i \geq 0$.
%The dataset is also a collection of pixel intensities $\{D_i\}_{i=1}^n$.
%The prior for the dataset given the parameters is a normal distribution
%with mean given by the underlying image $\{xi\}$ convolved by
%a point spread function. The standard deviation of each pixel is
%$\sigma$, assumed known. This is usually written as

%\begin{eqnarray}
%\boldsymbol{D} = \boldsymbol{M}\boldsymbol{x} + \boldsymbol{n}
%\end{eqnarray}
%where $\boldsymbol{n}$ is the noise in each pixel and $\boldsymbol{M}$ is
%the blurring operator written in matrix form.

%The prior for the pixel intensities $\{x_i\}$ has an ``entropic'' form,
%where images are upweighted (relative to a uniform prior) if the pixel
%intensities have a low dynamic range. The prior density is defined conditional
%on a hyperparameter $\lambda$:
%\begin{eqnarray}
%p(\boldsymbol{x} | \lambda) &=&
%\frac{\exp\left[\lambda S(\xx)\right]}{\int \exp\left[\lambda S(\xx)\right] \, d\xx}
%\label{eq:entropy_prior}
%\end{eqnarray}
%where
%\begin{eqnarray}
%S(\xx) &=& -\sum_{i=1}^n x_i \ln x_i
%\end{eqnarray}
%is the ``configurational entropy'' of the image (crucially, this is a function of
%an image, and not a function of a probability distribution).
%The denominator in Equation~\ref{eq:entropy_prior} ensures the conditional prior
%is normalized.

%In the 1980s, this prior was thought to have a fundamental status, but this
%is no longer the case \citep{massinf}, although the function $S(\xx)$ does
%have special properties when its argument is a measure \citep{knuth_skilling}.
%However, it is still a prior whose
%consequences we might be interested in.

%Instead, we can consider the configurational
%entropy $S(\xx)$ as simply a function of an image that may
%be relevant somehow, and that we are uncertain about. The joint posterior
%for $\xx$ and $\lambda$ is:
%\begin{eqnarray}
%p(\lambda, \xx | \data) &=&
%\frac{p(\lambda)p(\xx | \lambda)p(\data | \xx)}
%{p(\data)}\\
%&=&
%\frac{p(\lambda)p(\xx | \lambda)p(\data | \xx)}
%{p(\data)}
%\end{eqnarray}

%The posterior for the underlying image given the data is
%\begin{eqnarray}
%p(\boldsymbol{x} | \boldsymbol{D}, \lambda)
%&=& \frac{p(\boldsymbol{x} | \lambda)p(\boldsymbol{D}|\boldsymbol{x}, \lambda)}
%{\int p(\boldsymbol{x}|\lambda)p(\boldsymbol{D}|\boldsymbol{x}, \lambda) \,d^n\boldsymbol{m}}\\
%&=& \frac{\exp\left[-\lambda \sum_i x_i\ln x_i
%-\frac{1}{2\sigma^2}\left(D_i - \sum_{j} P_{ij}x_{j}\right)^2\right]}
%{\int \exp\left[-\lambda \sum_i x_i\ln x_i
%-\frac{1}{2\sigma^2}\left(D_i - \sum_{j} P_{ij}x_{j}\right)^2\right]\, d^n\boldsymbol{m}}
%\end{eqnarray}

%This is a `doubly intractable' problem \citep{murray}, since an integral, whose
%result is a function of the parameters, appears in the {\em denominator} of the
%target density. However, in our case the integral in the
%denominator arises from the prior for the parameters, rather than from the
%likelihood function.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{BJB acknowledges research and study leave funding from the
University of Auckland. We would like to thank Gábor Csányi,
Livia Bartók-Pártay,
Rob Salomone, John Skilling, etc etc.}
